Vagueness, Noise, and Signaling

### 2015-05-15 paper/version_03/paper.tex
revision version 3; this version sent

A number of authors have recently tried to explain why vagueness evolved as something that is
itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing or particular learning strategies
\citep[e.g.][]{FrankeJager2010:Vagueness-Signa,OConnor2013:The-Evolution-o}. This paper
contributes to the latter line of thought. Concretely, we introduce the replicator diffusion
dynamics that integrates stochastic noise on the differential confusability of similar
stimuli. The dynamic proposed here generalizes and complements previous like-minded
accounts. We show that stochastic noise can not only lead to vague, yet relatively efficient
signal use, but can also unify evolutionary outcomes and help avoid suboptimal categorization.

The next section introduces the background against which the work presented here can be
appreciated. Section~\ref{sec:repl-diff-dynam} introduces the replicator diffusion dynamic,
which is then explored in Section~\ref{sec:exploring-rdd}.  Section~\ref{sec:discussion}
reflects and compares our approach to others. Two appendices cover technical material that is
not strictly necessary to appreciate the main conceptual points of the paper.
Appendix~\ref{sec:repl-mutat-dynam} shows that out replicator diffusion dynamic is a special
case of the well-known replicator mutator dynamics. Appendix~\ref{sec:deriv-rdd-imit} gives a
derivation of our dynamic as the most-likely path of a population whose agents imitate
successful behavior without being aware of perceptual noise themselves.

@@

A sim-max game consists of a set of states $\States$, a set of messages $\Messgs$ with much
fewer messages than states, a prior probability distribution $\Pr \in \Delta(\States)$ such
that $\Pr(\state)$ gives the probability that state $\state$ occurs, a similarity
metric\footnote{A metric is a function of distance between any two points in a given space. It
  should satisfy certain axioms that one would intuitively expect from the words ``similarity''
  and ``distance'' alone, but these details do not matter for the purposes of this paper.}
  % : similarity should always be non-negative, zero only for the distance
  % between an element and itself, symmetric, and satisfy the triangle inequality that
  % $\similarity(x,z) \le \similarity(x,y) + \similarity(y,z)$
on states $\similarity \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\similarity(\mystate{1}, \mystate{2})$ is the (perceptual) similarity between $\mystate{1}$
and $\mystate{2}$, and a utility function $\Utils \mycolon \States \times \States \rightarrow
\mathds{R}$ such that $\Utils(\mystate{1}, \mystate{2})$ is the payoff for sender and receiver
for a play with actual state $\mystate{1}$ and receiver interpretation $\mystate{2}$. We
identify the receiver's acts with the states of the world, so that the game is one of guessing
the actual state, so to speak. The utility function should be a monotonically decreasing
function of similarity.

\citet{JagerMetzger2011:Voronoi-Languag} showed that the
evolutionarily stable states of sim-max games are remarkably
systematic. Their results were obtained for games with infinitely many
states in $n$-dimensional Euclidean space $\States \subseteq
\mathds{R}^n$ and a quadratic loss function for utilities
$\Utils(\mystate{1}, \mystate{2}) = - (\mystate{1} -
\mystate{2})^2$. For these games, the evolutionarily stable states are
demonstrably so-called Voronoi languages. Roughly put, a Voronoi
language is a pair of sender and receiver strategies, such that the
sender strategy partitions the state space into convex categories,
while the receiver's interpretations are the central spots in each
category. A subset $X$ of $\mathds{R}^n$ is convex if, informally put,
all points in $X$ are connected via a straight line that lies entirely
in $X$; $X$ has no gaps or dents. For example, if \States is the unit
interval and all states are equiprobable, a Voronoi language with two
messages would have the sender use one message exclusively for all
points in the lower half of the unit interval and another for all
points in the upper half; the receiver's interpretations of messages
are the central points, .25 and .75, in the respective intervals.
See Figure~\ref{fig:Voronoi-language-example} for an illustration.

@@

This result is interesting, because it demonstrates that signaling can
impose orderly categories on a metric space, without that being the
ulterior purpose of it all. Sender and receiver can be distinct
entities, whose purpose is to communicate effectively about the actual
state. In that case, evolving Voronoi languages would explain why
\emph{linguistic categories} are well-behaved and orderly in the way
they appear to be. More abstractly, sender and receiver can also be
thought of as distinct modules in a single system, where the first
module must discretize the information it is fed by selecting a small
sample of, suggestively, category labels. These are passed to a second
module that tries to decode the original information. In this case,
evolving Voronoi languages would explain why \emph{conceptual
  categories} are well-behaved and orderly in the way that they appear
to be.

@@

An alternative approach is taken by
\citet{FrankeJager2010:Vagueness-Signa} and
\citet{OConnor2013:The-Evolution-o} who show how the above desiderata
can be met by evolving strategies in sim-max games. The approach we
take here is similar in spirit, but different in relevant detail. We
argue here that \citeauthor{FrankeJager2010:Vagueness-Signa}'s
approach gives rise to problematic predictions under certain
conditions which are avoided in our approach. While
\citeauthor{OConnor2013:The-Evolution-o} gives an agent-level learning
dynamic under which vague signaling arises, we complement the picture
by giving a population-level dynamic that abstracts from the concrete
behavior of signalers. A more in-depth comparison is deferred until
Section~\ref{sec:discussion}. Let us first look into our own proposal
in more detail.

@@

We introduce a population-level dynamic in which communicatively more
efficient strategies increase in proportion, as usual, but where the
realization of signaling strategies is noisy because of the agents'
inability to sharply distinguish similar states. We call our dynamic
\emph{replicator diffusion dynamic} (\rdd), because it is an extension
of the replicator dynamic (\rd) and a special case of the replicator
mutator dynamic (\rmd).

The formulation of the \rdd given and explored here is in terms of behavioral strategies, not
in terms of mixed strategies. Dynamics on behavioral strategies assume that agents can adapt
their behavior locally, so to speak, i.e., independently at each choice
point. Section~\ref{sec:repl-mutat-dynam} elaborates on the
difference. Section~\ref{sec:relating-dynamics} shows how to translate our initial formulation
of the \rdd into one using mixed strategies. Doing so identifies the \rdd as a special case of
the \rmd.

Our focus on behavioral strategies greatly reduces the complexity of the dynamic and simplifies
numerical simulations. But it also points towards an interpretation of the process as cultural
evolution, rather than biological evolution. Signaling strategies are not inherited as a whole,
but adapted piece-meal, e.g., by conditional imitation of ``local'' behavior at a given choice
point. Appendix~\ref{sec:deriv-rdd-imit} gives one possible derivation of our dynamic as the
most likely path in a population of agents that imitate successful behavior while being
entirely unaware of perceptual imprecision themselves. Still, the possibility to formulate the
\rdd as a special case of the \rmd (Appendix~\ref{sec:diffusion-as-special}) also suggests that
an interpretation as a process of biological evolution may be feasible.

For now, let us first focus on the formulation of the dynamic in terms of behavioral strategies
alone.  We begin by recapitulating the \rd in Section~\ref{sec:repl-dynam-behav}, and introduce
the \rdd in Section~\ref{sec:repl-diff-dynam-1}.
%Section~\ref{sec:repl-mutat-dynam} introduces the \rmd, and
%Section~\ref{sec:diffusion-as-special} shows in what sense the \rdd is
%a special case of the \rmd. Sections~\ref{sec:repl-mutat-dynam} and
%\ref{sec:diffusion-as-special} are conceptually important, but rather
%technical and not strictly necessary to appreciate the main point of
%the paper.

@@

The \emph{fitness} of a behavioral strategy at a choice point is the
frequency-weighted average of expected utilities of each choice, given
the opponent's strategy:
\begin{align*}
  \Phi(\state,\Sstrat, \Rstrat) & = \sum_{\messg} \Sstrat(\messg \probbar \state) \cdot
\EU(\messg, \state,\Rstrat) \\
\Phi(\messg,\Sstrat, \Rstrat) & = \sum_{\act} \Rstrat(\act \probbar \messg)
\cdot \EU(\act, \messg,\Sstrat)\,.
\end{align*}
The discrete-time replicator dynamic maps current strategies $\Sstrat$
and $\Rstrat$ to future strategies $\RD(\Sstrat)$ and $\RD(\Rstrat)$
in such a way that changes in frequency are proportional to expected
utilities. For behavioral strategies, the changes take place locally
at each of the agents' choice points:
\begin{align*}
  \RD(\Sstrat)(\messg \probbar \state) & = \frac{\Sstrat(\messg \probbar \state) \cdot
    \EU(\messg, \state,\Rstrat)} {\Phi(\state,\Sstrat, \Rstrat)} \\
    \RD(\Rstrat)(\act \probbar \messg) & = \frac{\Rstrat(\act \probbar \messg) \cdot
    \EU(\act, \messg,\Sstrat)} {\Phi(\messg,\Sstrat, \Rstrat)}  \,.
\end{align*}
The \rd in behavioral strategies describes the most likely path of development of average
behavior in a population of agents when behavioral adaptations occur independently at each
choice point, e.g., when agents adopt the behavior of another agent \emph{at a given choice
  point} if that ``local'' behavior is ``locally'' better than their own (see
\cite{Sandholm2010:Population-Game} for agent-level interpretations of the replicator dynamics
and \citeauthor{Cressman2003:Evolutionary-Dy} \citeyear{Cressman2003:Evolutionary-Dy} for
applications to extensive form games and behavioral strategies).

@@

The aggregate effect of confusion of states on behavioral strategies
can be captured in a function that maps behavioral strategies
$\Sstrat$ and $\Rstrat$ to their \emph{diffusions} $D(\Sstrat)$ and
$D(\Rstrat)$. Suppose that $\Sstrat$ and $\Rstrat$ represent a
population's average signaling behavior if there was no confusion of
states. Then $D(\Sstrat)$ and $D(\Rstrat)$ give the expected aggregate
population behavior under the actual noise-perturbed behavior of each
agent. Since the effect of confusion of states is that behavior at one
choice point percolates to behavior at similar choice points, we speak
of \emph{diffusion of behavior under confusion of states}. If we
conceive of behavioral strategies $\Sstrat$ and $\Rstrat$ as
row-stochastic matrices (rows are choice points, columns are choices;
each row is a probability distribution), we can define the diffusion effect of
confusion matrix $C$ as matrix multiplication:
\begin{align}
  \label{eq:confusion-function}
  \Diff_C(\Sstrat) & = C \Sstrat &    \Diff_C(\Rstrat) & = \Rstrat C\,.
\end{align}
For example, diffusion thus defined determines that, on the diffused sender strategy, the probability that message $\messg$ is chosen given a state $\state_i$ is given by $\sum_{j} C_{ij} \cdot \Sstrat(\messg \probbar \state_j)$.
This means that the original probability given by $\Sstrat$ is increased by the original probability of this message being chosen in other states, proportionally to how likely it is that they are mistaken for $\state_i$.

@@

The most straightforward interpretation of this sequential definition is that the \rdd captures
the most likely path of development of aggregate behavior in a society, in which behavioral
changes occur independently at each choice point, as described by the \rd for behavioral
strategies. Perceptual noise affects the realization of behavior, but not the fitness-based
selection thereof. Alternative noise-perturbed versions of the replicator dynamics in which the
fitness-based selection of strategies takes noise into account are conceivable, but we must put
the exploration of alternatives aside here. We suggest that our approach is practical,
conservative and plausible: it is practical because it is easy to implement; it is conservative
because it connects the \rdd with the \rmd (see Appendix~\ref{sec:diffusion-as-special}); it is
also plausible because the \rdd can be derived from a myopic update protocol in which agents
themselves are unaware of perceptual noise (see Appendix~\ref{sec:deriv-rdd-imit}).

@@

The \rdd is as a two-step procedure: first calculate where the replicator dynamic takes us,
then apply diffusion. To understand the \rdd better, we isolate the novel diffusion part in
Section~\ref{sec:iterated-diffusion}, before exploring how diffusion and fitness-based
selection interact in Section~\ref{sec:simulations}.

@@

Now that we understand better what diffusion does, let us look at how
diffusion interacts with optimization of behavior, as described by the
replicator dynamics. The main question we should address is whether
diffusion and fitness-based replication reasonably interact, and if
so, whether the inclusion of diffusion has any noteworthy effects on
the evolving meaning of signals. Clearly, if $\imprecision = 0$, the
\rdd reduces to the \rd. If $\imprecision \rightarrow \infty$, the
diffusion component takes over and the \rdd reduces to the trivial
iterated diffusion process that we looked at in the previous
section. We will show presently that for reasonable in-between levels
of imprecision $\imprecision$, the \rdd can lead to reasonably
efficient, yet vague signaling behavior. Suitable levels of
imprecision can have further accelerating and, surprisingly, unifying
effects on meaning evolution.

@@

The \rdd adds expected confusion of similar states to the \rd in
behavioral strategies. The result is an abstract evolutionary dynamic
that describes the most likely path of development of aggregate
population behavior in which any process of behavioral adaptation at
the agent level that is consistent with the \rd in behavioral
strategies is perturbed by systematic noise in the realization of the
agents' strategies. Given the use of behavioral strategies, an
interpretation as a process of cultural evolution is most salient,
according to which agents adapt their choice dispositions locally at
each choice point. But the \rdd is also extensible to biological
evolution on mixed strategies, in which case it is a special case of
the \rmd with mutation matrices derived from confusion matrices, as
shown in Appendix~\ref{sec:diffusion-as-special}.

@@

\citet{OConnor2013:The-Evolution-o} introduces a generalization of classical Herrnstein
reinforcement learning for sim-max games, in which successful play leads to reinforcement of
acts also for states similar to the one that actually occurred. This generalization not only
leads to vague signaling of the appropriate kind, but also speeds up learning in such a way
that, especially for sim-max games with higher numbers of states, higher levels of
communicative success are reached in shorter learning periods. % Technically, this result is
% partly due to the fact that signalers make bigger changes to their behavioral strategies
% after each round of play under generalized reinforcement than under the Herrnstein
% variety. That only explains the speed of adaptation, not necessarily that generalization also
% leads to regularity and communicative efficiency, but it does.

Diffusion of strategies in the \rdd can also be conceived of as a form
of generalization, and works in large part quite analogously to stimulus
generalization in \citeauthor{OConnor2013:The-Evolution-o}'s
approach. But there are still
differences. \citeauthor{OConnor2013:The-Evolution-o} showed that
generalized reinforcement learning can speed up the development of
efficient signaling, especially for higher numbers of
states. Complementing this, we showed that diffusion regularizes
evolutionary outcomes and prevents meandering around sub-optimal and
non-convex strategies, also for low numbers of states.

Conceptually, the \rdd is a more abstract framework than generalized
reinforcement learning. The latter is most naturally seen as an
agent-based learning dynamic that has two players adapt their
individual strategies after each concrete round of play. In contrast,
the \rdd describes a more abstract, average dynamical change in
behavioral dispositions at the aggregate population level. Although
the behavior of (some forms of) reinforcement learning mirror those of
the \rd at some stage in time
(\cite{BorgersSarin997:Learning-Throug,HopkinsPosch2005:Attainability-o,Beggs2005:On-the-Converge}),
this does not mean that \emph{generalized} reinforcement learning is
also necessarily a plausible population-level description of, say,
generalized learning in a population of agents. Seen in this light,
generalized reinforcement learning and the \rdd nicely complement each
other, as similarly-minded accounts operating at different levels of
abstraction.
Although it would be interesting to explore the possibility of a mathematical connection between the two, this is beyond the scope of this paper and thus left as future work.

\paragraph{Quantal response equilibria.}
\citet{FrankeJager2010:Vagueness-Signa} suggested a number of ways in which
information-processing limitations could lead to vague strategies. The model that is most
clearly related to the present approach uses the notion of a \emph{quantal response}, also
known as a \emph{soft-max function}
\citep[e.g.][]{Luce1959:Individual-Choi,McFadden1976:Quantal-Choice-,GoereeHolt2008:Quantal-Respons}. A
quantal response is a stochastic generalization of the best response function (i.e., the
deterministic choice of an action that maximizes expected utility). Quantal responses factor in
stochastic trembles at the level of the computation or maximization of expected utilities:
agents may make mistakes in calculating how good given choice options really are for
them. \citet{FrankeJager2010:Vagueness-Signa} show that quantal response equilibria of sim-max
games can show the desired marks of vague signaling. Figure~\ref{fig:exampleQRE_stratsA} gives
an example of a quantal response equilibrium for a sim-max game, as used in our set-up but with
higher tolerance $\toler = 0.5$. Sender and receiver strategies look very much like what
evolves under \rdd with modest values of perceptual imprecision.

@@

We introduced the replicator diffusion dynamic and identified it as a special case of the
replicator mutator dynamic (Appendix~\ref{sec:diffusion-as-special}). The \rdd incorporates a
special kind of stochastic mutation due to differential confusability of similar states. It can
be thought of as the aggregate population-level behavior process that results from imitation
among agents that are unaware of noise (Appendix~\ref{sec:deriv-rdd-imit}). Based on data from
numerical simulations, we demonstrated that the inclusion of mild levels of stochastic
confusability of states does not undermine the possibility of evolving relatively successful
signaling strategies. On the contrary, strategies that evolved under mild diffusion induce a
highly regular and systematic category structure that shows the signs of vagueness as
desired. There might even be a higher-order benefit to the presence of imprecision, in that it
can accelerate convergence to optimal categorization, by preventing evolutionary paths to stay
near inefficient non-convex strategies for a long time. Diffusion thus also unifies and
regularizes evolutionary outcomes in such a way that sub-optimal categorization is avoided in
short-term trajectories.

@@

The usual formulation of the \rmd is based on mixed, not behavioral
strategies. But a discrete-time formulation of the \rmd in behavioral
strategies clearly identifies the \rdd to be a special case. It
follows that a mixed strategy version of the \rdd is just the \rmd
with suitably defined mutation matrices. From this, even a
continuous-time version can easily be retrieved if desired. To show
this, we start with a delineation of mixed and behavioral strategies.

>>>>

A number of authors have recently tried to explain why vagueness evolved as something that is
itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing or particular learning strategies
\citep[e.g.][]{FrankeJager2010:Vagueness-Signa,OConnor2013:The-Evolution-o}. This paper
contributes to the latter line of thought. Concretely, we introduce the replicator diffusion
dynamic that integrates confusability of similar stimuli. The dynamic proposed here generalizes
and complements previous like-minded accounts. We show that stochastic noise can not only lead
to vague, yet relatively efficient signal use, but can also unify evolutionary outcomes and
help avoid suboptimal categorization.

The next section introduces the background against which the work presented here can be
appreciated. Section~\ref{sec:repl-diff-dynam} introduces the replicator diffusion dynamic,
which is then explored in Section~\ref{sec:exploring-rdd}.  Section~\ref{sec:discussion}
reflects and compares our approach to others. Two appendices cover technical material that is
not strictly necessary to appreciate the main conceptual points of the paper.
Appendix~\ref{sec:diffusion-as-special} shows that our replicator diffusion dynamic is a
special case of the well-known replicator mutator dynamic. Appendix~\ref{sec:deriv-rdd-imit}
gives a derivation of our dynamic as the most-likely path of a population whose agents imitate
successful behavior without being aware of perceptual noise themselves.

@@

A sim-max game consists of a set of states $\States$, a set of messages $\Messgs$ with much
fewer messages than states, a probability distribution $\Pr \in \Delta(\States)$ such that
$\Pr(\state)$ gives the probability that state $\state$ occurs, a similarity metric\footnote{A
  metric is a function of distance between any two points in a given space. It should satisfy
  certain axioms that ensure behavior that one would intuitively expect from the words
  ``similarity'' and ``distance'' alone, but these details do not matter for the purposes of
  this paper.}
  % : similarity should always be non-negative, zero only for the distance
  % between an element and itself, symmetric, and satisfy the triangle inequality that
  % $\similarity(x,z) \le \similarity(x,y) + \similarity(y,z)$
on states $\similarity \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\similarity(\mystate{1}, \mystate{2})$ is the (perceptual) similarity between $\mystate{1}$
and $\mystate{2}$, and a utility function $\Utils \mycolon \States \times \States \rightarrow
\mathds{R}$ such that $\Utils(\mystate{1}, \mystate{2})$ is the payoff for sender and receiver
for a play with actual state $\mystate{1}$ and receiver interpretation $\mystate{2}$. We
identify the receiver's acts with the states of the world, so that the game is one of guessing
the actual state, so to speak. The utility function should be a monotonically decreasing
function of similarity.

\citet{JagerMetzger2011:Voronoi-Languag} showed that the evolutionarily stable states of
sim-max games are remarkably systematic: the evolutionarily stable states are demonstrably
so-called Voronoi languages.\footnote{Their results were obtained for games with infinitely
  many states in $n$-dimensional Euclidean space $\States \subseteq \mathds{R}^n$ and a
  quadratic loss function for utilities
  $\Utils(\mystate{1}, \mystate{2}) = - (\mystate{1} - \mystate{2})^2$.} Roughly put, a Voronoi
language is a pair of sender and receiver strategies, such that the sender strategy partitions
the state space into convex categories, while the receiver's interpretations are the central
spots in each category. A subset $X$ of $\mathds{R}^n$ is convex if, informally put, all points
in $X$ are connected via a straight line that lies entirely in $X$; $X$ has no gaps or
dents. For example, if \States is the unit interval and all states are equiprobable, a Voronoi
language with two messages would have the sender use one message exclusively for all points in
the lower half of the unit interval and another for all points in the upper half; the
receiver's interpretations of messages are the central points, .25 and .75, in the respective
intervals.  See Figure~\ref{fig:Voronoi-language-example} for an illustration.

@@

This result is interesting because it demonstrates that signaling can
impose orderly categories on a metric space, without that being the
ulterior purpose of it all. Sender and receiver can be distinct
entities, whose purpose is to communicate effectively about the actual
state. In that case, evolving Voronoi languages would explain why
\emph{linguistic categories} are well-behaved and orderly in the way
they appear to be. More abstractly, sender and receiver can also be
thought of as distinct modules in a single system, where the first
module must discretize the information it is fed by selecting a small
sample of, suggestively, category labels. These are passed to a second
module that tries to decode the original information. In this case,
evolving Voronoi languages would explain why \emph{conceptual
  categories} are well-behaved and orderly in the way that they appear
to be.

@@

An alternative approach is taken by \citet{FrankeJager2010:Vagueness-Signa} and
\citet{OConnor2013:The-Evolution-o} who show how the above desiderata can be met by evolving
strategies in sim-max games. The approach we take here is similar in spirit, but different in
relevant detail. A more in-depth comparison is deferred until Section~\ref{sec:discussion}. Let
us first look into our own proposal in more detail.

@@

We introduce a population-level dynamic in which communicatively more
efficient strategies increase in proportion, as usual, but where the
realization of signaling strategies is noisy because of the agents'
inability to sharply distinguish similar states. We call our dynamic
\emph{replicator diffusion dynamic} (\rdd) because it is an extension
of the replicator dynamic (\rd) and a special case of the replicator
mutator dynamic (\rmd).

The formulation of the \rdd given and explored here is in terms of behavioral strategies, not
in terms of mixed strategies. Dynamics on behavioral strategies assume that agents can adapt
their behavior locally, so to speak, i.e., independently at each choice
point. Appendix~\ref{sec:repl-mutat-dynam} elaborates on the
difference. Appendix~\ref{sec:relating-dynamics} shows how to translate our initial formulation
of the \rdd into one using mixed strategies. Doing so identifies the \rdd as a special case of
the \rmd.

Our focus on behavioral strategies greatly reduces the complexity of the dynamic and simplifies
numerical simulations. But it also points towards an interpretation of the process as cultural
evolution, rather than biological evolution. Signaling strategies are not inherited as a whole,
but adapted piece-meal, e.g., by conditional imitation of ``local'' behavior at a given choice
point. Appendix~\ref{sec:deriv-rdd-imit} gives one possible derivation of our dynamic as the
most likely path of development in a population of agents that imitate successful behavior
while being entirely unaware of perceptual imprecision themselves. Still, the possibility to
formulate the \rdd as a special case of the \rmd (Appendix~\ref{sec:diffusion-as-special}) also
suggests that an interpretation as a process of biological evolution may be feasible.

We begin by recapitulating the \rd in Section~\ref{sec:repl-dynam-behav}, and introduce
the \rdd in Section~\ref{sec:repl-diff-dynam-1}.
%Section~\ref{sec:repl-mutat-dynam} introduces the \rmd, and
%Section~\ref{sec:diffusion-as-special} shows in what sense the \rdd is
%a special case of the \rmd. Sections~\ref{sec:repl-mutat-dynam} and
%\ref{sec:diffusion-as-special} are conceptually important, but rather
%technical and not strictly necessary to appreciate the main point of
%the paper.

@@

The \emph{fitness} of a behavioral strategy at a choice point is the
frequency-weighted average of expected utilities of each choice, given
the opponent's strategy:
\begin{align*}
  \Phi(\state,\Sstrat, \Rstrat) & = \sum_{\messg} \Sstrat(\messg \probbar \state) \cdot
\EU(\messg, \state,\Rstrat) \\
\Phi(\messg,\Sstrat, \Rstrat) & = \sum_{\act} \Rstrat(\act \probbar \messg)
\cdot \EU(\act, \messg,\Sstrat)\,.
\end{align*}
The discrete-time replicator dynamic maps current strategies $\Sstrat$
and $\Rstrat$ to future strategies $\RD(\Sstrat)$ and $\RD(\Rstrat)$
in such a way that changes in frequency are proportional to expected
utilities. For behavioral strategies, the changes take place locally
at each of the agents' choice points:
\begin{align*}
  \RD(\Sstrat)(\messg \probbar \state) & = \frac{\Sstrat(\messg \probbar \state) \cdot
    \EU(\messg, \state,\Rstrat)} {\Phi(\state,\Sstrat, \Rstrat)} \\
    \RD(\Rstrat)(\act \probbar \messg) & = \frac{\Rstrat(\act \probbar \messg) \cdot
    \EU(\act, \messg,\Sstrat)} {\Phi(\messg,\Sstrat, \Rstrat)}  \,.
\end{align*}
The \rd in behavioral strategies describes the most likely path of development of average
behavior in a population of agents when behavioral adaptations occur independently at each
choice point, e.g., when agents adopt the behavior of another agent \emph{at a given choice
  point} if that ``local'' behavior is ``locally'' better than their own (see
\cite{Sandholm2010:Population-Game} for agent-level interpretations of the replicator dynamic
and \citeauthor{Cressman2003:Evolutionary-Dy} \citeyear{Cressman2003:Evolutionary-Dy} for
applications to extensive form games and behavioral strategies).

@@

The aggregate effect of confusion of states on behavioral strategies
can be captured in a function that maps behavioral strategies
$\Sstrat$ and $\Rstrat$ to their \emph{diffusions} $D(\Sstrat)$ and
$D(\Rstrat)$. Suppose that $\Sstrat$ and $\Rstrat$ represent a
population's average signaling behavior if there was no confusion of
states. Then $D(\Sstrat)$ and $D(\Rstrat)$ give the expected aggregate
population behavior under the actual noise-perturbed behavior of each
agent. Since the effect of confusion of states is that behavior at one
choice point percolates to behavior at similar choice points, we speak
of \emph{diffusion of behavior under confusion of states}. If we
conceive of behavioral strategies $\Sstrat$ and $\Rstrat$ as
row-stochastic matrices (rows are choice points, columns are choices;
each row is a probability distribution), we can define the diffusion effect of
confusion matrix $C$ as matrix multiplication:
\begin{align}
  \label{eq:confusion-function}
  \Diff_C(\Sstrat) & = C \Sstrat &    \Diff_C(\Rstrat) & = \Rstrat C\,.
\end{align}
For example, diffusion thus defined determines that, on the diffused sender strategy, the
probability that message $\messg$ is chosen given a state $\mystate{i}$ is given by
$\sum_{j} C_{ij} \cdot \Sstrat(\messg \probbar \mystate{j})$.  This means that the original
probability given by $\Sstrat$ is increased by the original probability of this message being
chosen in other states $\mystate{j}$, proportionally to how likely it is that $\mystate{i}$ is
(mistakenly) perceived as $\mystate{j}$.

@@

The most straightforward interpretation of this sequential definition is that the \rdd captures
the most likely path of development of aggregate behavior in a society, in which behavioral
changes occur independently at each choice point, as described by the \rd for behavioral
strategies. Perceptual noise affects the realization of behavior, but not the fitness-based
selection thereof. Alternative noise-perturbed versions of the replicator dynamic in which the
fitness-based selection of strategies takes noise into account are conceivable, but we must put
the exploration of alternatives aside here. We suggest that our approach is practical,
conservative and plausible: it is practical because it is easy to implement; it is conservative
because it connects the \rdd with the \rmd (see Appendix~\ref{sec:diffusion-as-special}); it is
also plausible because the \rdd can be derived from a myopic update protocol in which agents
themselves are unaware of perceptual noise (see Appendix~\ref{sec:deriv-rdd-imit}).

@@

The \rdd is a two-step procedure: first calculate where the replicator dynamic takes us, then
apply diffusion. To understand the \rdd better, we isolate the novel diffusion part in
Section~\ref{sec:iterated-diffusion}, before exploring how diffusion and fitness-based
selection interact in Section~\ref{sec:simulations}.

@@

Now that we understand better what diffusion does, let us look at how
diffusion interacts with optimization of behavior, as described by the
replicator dynamic. The main question we should address is whether
diffusion and fitness-based replication reasonably interact, and if
so, whether the inclusion of diffusion has any noteworthy effects on
the evolving meaning of signals. Clearly, if $\imprecision = 0$, the
\rdd reduces to the \rd. If $\imprecision \rightarrow \infty$, the
diffusion component takes over and the \rdd reduces to the trivial
iterated diffusion process that we looked at in the previous
section. We will show presently that for reasonable in-between levels
of imprecision $\imprecision$, the \rdd can lead to reasonably
efficient, yet vague signaling behavior. Suitable levels of
imprecision can have further accelerating and, surprisingly, unifying
effects on meaning evolution.

@@

The \rdd adds expected confusion of similar states to the \rd in behavioral strategies. The
result is an abstract evolutionary dynamic that describes the most likely path of development
of aggregate population behavior that is consistent with the idea that agents imitate other
agents' behavior (Appendix~\ref{sec:deriv-rdd-imit}). This would be a process of cultural
evolution operating on behavioral strategies. But we speculate that the \rdd is also extensible
to biological evolution on mixed strategies, in which case it is a special case of the \rmd
with mutation matrices derived from confusion matrices, as shown in
Appendix~\ref{sec:diffusion-as-special}.

@@

\citet{OConnor2013:The-Evolution-o} introduces a variant of classical Herrnstein reinforcement
learning for sim-max games, in which successful play leads to reinforcement of acts also for
states similar to the one that actually occurred. This is a form of generalization to unseen
contingencies. It not only leads to vague signaling of the appropriate kind, but also speeds up
learning in such a way that, especially for sim-max games with higher numbers of states, higher
levels of communicative success are reached in shorter learning
periods. % Technically, this result is
% partly due to the fact that signalers make bigger changes to their behavioral strategies
% after each round of play under generalized reinforcement than under the Herrnstein
% variety. That only explains the speed of adaptation, not necessarily that generalization also
% leads to regularity and communicative efficiency, but it does.

Diffusion of strategies in the \rdd can also be conceived of as a form of generalization, and
works in large part quite analogously to stimulus generalization in
\citeauthor{OConnor2013:The-Evolution-o}'s approach. But there are still
differences. \citeauthor{OConnor2013:The-Evolution-o} showed that her generalized reinforcement
learning can speed up the development of efficient signaling, especially for higher numbers of
states. Complementing this, we showed that diffusion regularizes evolutionary outcomes and
prevents meandering around sub-optimal and non-convex strategies, also for low numbers of
states.

Conceptually, the \rdd is a more abstract framework than generalized reinforcement
learning. The latter is most naturally seen as an agent-based learning dynamic that has two
players adapt their individual strategies after each concrete round of play. In contrast, the
\rdd describes a more abstract, average dynamical change in behavioral dispositions at the
aggregate population level. Although the behavior of (some forms of) reinforcement learning
mirror those of the \rd at some stage in time
(\cite{BorgersSarin997:Learning-Throug,HopkinsPosch2005:Attainability-o,Beggs2005:On-the-Converge}),
this does not mean that \emph{generalized} reinforcement learning is also necessarily a
plausible population-level description of, say, generalized learning in a population of
agents. Seen in this light, generalized reinforcement learning and the \rdd nicely complement
each other, as similarly-minded accounts operating at different levels of abstraction.
Although it would be interesting to explore the possibility of a mathematical connection
between the two, this is beyond the scope of this paper.

\paragraph{Quantal response equilibria.}
\citet{FrankeJager2010:Vagueness-Signa} suggested a number of ways in which
information-processing limitations could lead to vague strategies. The model that is most
clearly related to the present approach uses the notion of a \emph{quantal response}, also
known as a \emph{soft-max function}
\citep[e.g.][]{Luce1959:Individual-Choi,McFadden1976:Quantal-Choice-,GoereeHolt2008:Quantal-Respons}. A
quantal response is a stochastic generalization of the best response function (i.e., the
deterministic choice of an action that maximizes expected utility). Quantal responses factor in
stochastic trembles at the level of the computation or maximization of expected utilities:
agents may make mistakes in calculating how good given choice options really are for
them. \citet{FrankeJager2010:Vagueness-Signa} demonstrate that quantal response equilibria of
sim-max games can show the desired marks of vague
signaling. Figure~\ref{fig:exampleQRE_stratsA} gives an example of a quantal response
equilibrium for a sim-max game, as used in our set-up but with higher tolerance $\toler =
0.5$.
Sender and receiver strategies look very much like what evolves under \rdd with modest values
of perceptual imprecision.

@@

We introduced the replicator diffusion dynamic and identified it as a special case of the
replicator mutator dynamic (Appendix~\ref{sec:diffusion-as-special}). The \rdd incorporates a
special kind of stochastic mutation due to differential confusability of similar states. It can
be thought of as the aggregate population-level behavior process that results from imitation
among agents that are unaware of perceptual confusion
(Appendix~\ref{sec:deriv-rdd-imit}). Based on data from numerical simulations, we demonstrated
that the inclusion of mild levels of perceptual imprecision does not undermine the possibility
of evolving relatively successful signaling strategies. On the contrary, strategies that
evolved under mild imprecision induce a highly regular and systematic category structure that
shows the signs of vagueness as desired. There might even be a higher-order benefit to the
presence of imprecision, in that it can accelerate convergence to optimal categorization, by
preventing evolutionary paths to stay near inefficient non-convex strategies for a long time. A
little imprecision thus also unifies and regularizes evolutionary outcomes in such a way that
sub-optimal categorization is avoided in short-term trajectories.

@@

The usual formulation of the \rmd is based on mixed, not behavioral strategies. But a
discrete-time formulation of the \rmd in behavioral strategies clearly identifies the \rdd to
be a special case. It follows that a mixed-strategy version of the \rdd is just the \rmd with
suitably defined mutation matrices. From this, even a continuous-time version can easily be
retrieved if desired. To show this, we start with a delineation of mixed and behavioral
strategies.
